Type Classes for Mathematics in Type Theory

### 2010-06-30 jar.tex
Fixed scare quotes

Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully ``unbundled'' predicate-representations of algebraic structures practical to work with.

@@

\paragraph{Outline}
In section~\ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes). Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of ``unbundled'' purely predicate based interfaces for abstract structures practical. Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes. In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers. In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}. In section~\ref{sequences} we hint at an interface for sequences but describe how a limitation in the Coq system makes its use problematic. We end with conclusions in section~\ref{conclusions}.

@@

Occasionally there is some ambiguity as to whether a certain piece of information (such as a witness to an existential statement) is strictly ``proof matter'' (and thus belongs in the \lstinline|Prop| sort) or actually of further computational interest (and thus does \emph{not} belong to the \lstinline|Prop| sort). We will see one such case when we discuss the first homomorphism theorem in section \ref{homothm}. Coq provides a modest level of \emph{universe-polymorphism} so that we may avoid duplication when trying to support \lstinline|Prop|-sorted and non-\lstinline|Prop|-sorted content with a single set of definitions.

@@

The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve ``\lstinline|=|'' for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.

@@

Because the Coq type theory lacks ``real'' quotient types (as it would make type checking undecidable), the conventional way of working with abstract structures in Coq is to always start with \emph{setoids} (aka Bishop sets): types equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}. As we will see in section ~\ref{univ}, this way of working by nature lends itself well to particularly simple definitions of notions such as quotient algebras.

@@

In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus. By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (``dictionaries''), instances are ordinary constants of these record types (registered as ``hints'' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.

@@

In our example, lack of sharing support rears its head as soon as we try to define \lstinline|EquivalenceRelation| in terms of \lstinline|ReflexiveRelation| and its hypothetical siblings bundling symmetric and transitive relations. There, we would need some way to make sure that when we ``inherit'' \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation| by adding them as fields in our bundled record, they all refer to the same carrier and relation. Adding additional fields stating equalities between the three bundled carriers and relations is neither easily accomplished (because one would need to work with heterogenous equality) nor would it permit natural use of the resulting structure (because one would constantly have to rewrite things back and forth).

@@

The observant reader may wonder whether it might be beneficial to go one step further and unbundle proofs of laws and inherited substructures as well. This is not the case, because there is no point in sharing them. After all, by (weak) proof irrelevance, the ``value'' of such proofs can be of no consequence anyway. Indeed, parameterizing on proofs would be actively harmful because instantiations differing only in the proof argument would express the same thing yet be nonconvertible, requiring awkward conversions and hindering automation.

@@

As an aside, we note that the distinction between the class field name and the infix operator notation bound to it is really just a mildly awkward Coq artifact. In Haskell, where operators can themselves be used as names, there would be no need to have the \lstinline|equiv| and \lstinline|sg_op| names in addition to the operator ``names''.

@@

One aspect of the predicate approach we have not mentioned thus far is that in proofs parameterized by abstract structures, all components become hypotheses in the context. For the example theorem above, the context looks like:
\begin{lstlisting}
  G: Type
  e: Equiv G
  op: SemiGroupOp G
  P: SemiGroup G e op
\end{lstlisting}
We are not particularly worried about overly large contexts, especially because most of the ``extra'' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense ``inert'' with respect to proof search. Hence, we do not foresee problems with large contexts for any but the most complex formalizations.

@@

Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive). In our hierarchy, these ``substructures'' by themselves establish the following inheritance diagram:

@@

Since \lstinline|CommutativeMonoid| indirectly includes a \lstinline|SemiGroup| field, which in turn includes a \lstinline|Equivalence| field, having a \lstinline|SemiRing| proof means having two distinct proofs that the equality relation is an equivalence. This kind of redundant knowledge (which arises naturally in mathematics) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs. We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the ``right'' proofs of properties. There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than ``mere'' convertibility) to recognize identical subterms.

@@

The reader will note that we use names for properties like distributivity and absorption. Indeed, these are type classes as well (which is why we declare them with \lstinline|:>|). It has been our experience that almost any generic predicate worth naming is worth representing as a predicate type class, so that its proofs will be resolved as instances behind the scenes whenever possible. Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular. Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a ``database'' of properties previously proved. Much more so than the existing canonical structures facility, because canonical structures can only be used with bundled representations, which we argued against in section~\ref{bundling}.

@@

Initiality, too, is defined by a combination of an operational and a predicate class:
\begin{lstlisting}
  Context `{Category X}.
  Class InitialArrows (x: X): Type := initial_arrow: Π y, x ⟶ y.
  Class Initial (x: X) `{InitialArrows x}: Prop :=
    initial_arrow_unique: Π y (a: x ⟶ y), a = initial_arrow y.
\end{lstlisting}
The operational class \lstinline|InitialArrows| designates the arrows that originate from an initial object \lstinline|x| by virtue of it being initial. The \lstinline|Initial| class itself further requires these ``initial arrows'' to be unique. Having \lstinline|InitialArrows| as an operational type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where ``known'' should be read as ``can be determined by instance resolution'').

@@

  Class Functor `{Fmap}: Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎$$ g) = fmap f ◎$$ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: ``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.

@@

We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two distinct and interdependent maps is an abuse of notation of sufficient severity to make it ill-suited to a formal development where software engineering concerns apply. Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.

@@

To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed. While one could define both of them ``manually'', greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically (varieties are categories consisting of models for a fixed theory with homomorphisms between them).

@@

A multisorted signature enumerates sorts, operations, and specifies the ``types'' of the operations as non-empty lists of sorts, where the final element denotes the result type:
\begin{lstlisting}
  Inductive Signature: Type :=
    { sorts: Set
    ; operation:> Set
    ; operation_type:> operation → ne_list sorts }.
\end{lstlisting}
Given an interpretation of the sorts (mapping each symbolic sort to a carrier), an interpretation for the operations is easily written as an operational type class:
\begin{lstlisting}
  Variables (σ: Signature) (carriers: sorts σ → Type).

@@

What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the ``variety''). Since we need a type for the objects in the \lstinline|Category|, at this point we have no choice but to bundle components and proof together in a record:
\begin{lstlisting}
  Variable et: EquationalTheory.

@@

\lstinline|EquationalTheory|'s for semirings and rings are easy to define, and so from section~\ref{varieties} we get corresponding categories in which we can postulate initial objects:
\begin{lstlisting}
  Class Naturals (A: ObjectInVariety semiring_theory) `{!InitialArrow A}: Prop :=
    { naturals_initial:> Initial A }.
\end{lstlisting}
While succinct, this definition is not a satisfactory abstraction because the use of \lstinline|ObjectInVariety| for the type of the \lstinline|A| component ``leaks'' the fact that we used this one particular universal algebraic construction of the category, which is just an implementation choice. Furthermore, this definition needs an additional layer of class instances to relate it to the \lstinline|SemiRing| class from our algebraic hierarchy.

@@

Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then ``lifted'' to the abstract interface. For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations. For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using the fact that \lstinline|naturals_to_semiring| is an involution.

@@

Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this ``quoting''. In~\cite{Hints}, Asperti et. al. show that unification hints, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.

@@

With these, we can define the primary ingredient, the \lstinline|Quote| class:
\begin{lstlisting}
  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V'): Type :=
    { quote: Expr (V + V')
    ; eval_quote: eval (merge l r) quote = n }.
\end{lstlisting}
We can think of \lstinline|Quote| as the type for a family of Prolog-like syntax-directed resolution functions, which will take as input \lstinline|V| and \lstinline|l| representing previously encountered holes (opaque subexpressions that could not be destructured further) and their values, along with a concrete term \lstinline|n| to be quoted. Their ``output'' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values. Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting \lstinline|y| with existing heap \lstinline|x| generates new heap \lstinline|z|''.

@@

  Class Extend := extend: Π {x y} `{SemiGroupOp y} `{MonoidUnit y},
    (x → y) → (seq x → y).
  Class Singleton := singleton: Π x, x → seq x.
\end{lstlisting}
With these, we can define the predicate class for sequences:
\begin{lstlisting}
  Class Sequence
   `{Π a, MonoidUnit (seq a)} `{Π a, SemiGroupOp (seq a)}
   `{Π a, Equiv a → Equiv (seq a)} `{Singleton} `{Extend}: Prop := ...
\end{lstlisting}
%    { ... structural and propriety proofs sufficient to bundle up the categories and arrows ...
%    ; ... AltAdjunction stated using the bundled ... }.
On top of this interface we can build theory about typical list operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera. We can also generically define ``bigops'' for summation and products, and easily show properties like distributivity, all without ever mentioning cons lists.

>>>>

Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully `unbundled' predicate-representations of algebraic structures practical to work with.

@@

\paragraph{Outline}
In section~\ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes). Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of `unbundled' purely predicate based interfaces for abstract structures practical. Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes. In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers. In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}. In section~\ref{sequences} we hint at an interface for sequences but describe how a limitation in the Coq system makes its use problematic. We end with conclusions in section~\ref{conclusions}.

@@

Occasionally there is some ambiguity as to whether a certain piece of information (such as a witness to an existential statement) is strictly `proof matter' (and thus belongs in the \lstinline|Prop| sort) or actually of further computational interest (and thus does \emph{not} belong to the \lstinline|Prop| sort). We will see one such case when we discuss the first homomorphism theorem in section \ref{homothm}. Coq provides a modest level of \emph{universe-polymorphism} so that we may avoid duplication when trying to support \lstinline|Prop|-sorted and non-\lstinline|Prop|-sorted content with a single set of definitions.

@@

The `native' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.

@@

Because the Coq type theory lacks quotient types (as it would make type checking undecidable), the conventional way of working with abstract structures in Coq is to always start with \emph{setoids} (aka Bishop sets): types equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}. As we will see in section ~\ref{univ}, this way of working by nature lends itself well to particularly simple definitions of notions such as quotient algebras.

@@

In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus. By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (`dictionaries'), instances are ordinary constants of these record types (registered as `hints' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.

@@

In our example, lack of sharing support rears its head as soon as we try to define \lstinline|EquivalenceRelation| in terms of \lstinline|ReflexiveRelation| and its hypothetical siblings bundling symmetric and transitive relations. There, we would need some way to make sure that when we `inherit' \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation| by adding them as fields in our bundled record, they all refer to the same carrier and relation. Adding additional fields stating equalities between the three bundled carriers and relations is neither easily accomplished (because one would need to work with heterogenous equality) nor would it permit natural use of the resulting structure (because one would constantly have to rewrite things back and forth).

@@

The observant reader may wonder whether it might be beneficial to go one step further and unbundle proofs of laws and inherited substructures as well. This is not the case, because there is no point in sharing them. After all, by (weak) proof irrelevance, the `value' of such proofs can be of no consequence anyway. Indeed, parameterizing on proofs would be actively harmful because instantiations differing only in the proof argument would express the same thing yet be nonconvertible, requiring awkward conversions and hindering automation.

@@

As an aside, we note that the distinction between the class field name and the infix operator notation bound to it is really just a mildly awkward Coq artifact. In Haskell, where operators can themselves be used as names, there would be no need to have the \lstinline|equiv| and \lstinline|sg_op| names in addition to the operator `names'.

@@

One aspect of the predicate approach we have not mentioned thus far is that in proofs parameterized by abstract structures, all components become hypotheses in the context. For the example theorem above, the context looks like:
\begin{lstlisting}
  G: Type
  e: Equiv G
  op: SemiGroupOp G
  P: SemiGroup G e op
\end{lstlisting}
We are not particularly worried about overly large contexts, especially because most of the `extra' hypotheses we have compared to bundled approaches are declarations of relations, operators, and constants, which are all in some sense `inert' with respect to proof search. Hence, we do not foresee problems with large contexts for any but the most complex formalizations.

@@

Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive). In our hierarchy, these `substructures' by themselves establish the following inheritance diagram:

@@

Since \lstinline|CommutativeMonoid| indirectly includes a \lstinline|SemiGroup| field, which in turn includes a \lstinline|Equivalence| field, having a \lstinline|SemiRing| proof means having two distinct proofs that the equality relation is an equivalence. This kind of redundant knowledge (which arises naturally in mathematics) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs. We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the `right' proofs of properties. There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than `mere' convertibility) to recognize identical subterms.

@@

The reader will note that we use names for properties like distributivity and absorption. Indeed, these are type classes as well (which is why we declare them with \lstinline|:>|). It has been our experience that almost any generic predicate worth naming is worth representing as a predicate type class, so that its proofs will be resolved as instances behind the scenes whenever possible. Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular. Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a `database' of properties previously proved. Much more so than the existing canonical structures facility, because canonical structures can only be used with bundled representations, which we argued against in section~\ref{bundling}.

@@

Initiality, too, is defined by a combination of an operational and a predicate class:
\begin{lstlisting}
  Context `{Category X}.
  Class InitialArrows (x: X): Type := initial_arrow: Π y, x ⟶ y.
  Class Initial (x: X) `{InitialArrows x}: Prop :=
    initial_arrow_unique: Π y (a: x ⟶ y), a = initial_arrow y.
\end{lstlisting}
The operational class \lstinline|InitialArrows| designates the arrows that originate from an initial object \lstinline|x| by virtue of it being initial. The \lstinline|Initial| class itself further requires these ``initial arrows'' to be unique. Having \lstinline|InitialArrows| as an operational type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where `known' should be read as ``can be determined by instance resolution'').

@@

  Class Functor `{Fmap}: Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎$$ g) = fmap f ◎$$ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: \lstinline|F x| and \lstinline|F f| map an object and an arrow, respectively, because \lstinline|x| and \lstinline|f| are conventional names for objects and arrows, respectively.

@@

We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two distinct and interdependent maps is an abuse of notation of sufficient severity to make it ill-suited to a formal development where software engineering concerns apply. Hence, we do not adopt this practice, and use \lstinline|fmap F| (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.

@@

To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed. While one could define both of them manually, greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically (varieties are categories consisting of models for a fixed theory with homomorphisms between them).

@@

A multisorted signature enumerates sorts, operations, and specifies the `types' of the operations as non-empty lists of sorts, where the final element denotes the result type:
\begin{lstlisting}
  Inductive Signature: Type :=
    { sorts: Set
    ; operation:> Set
    ; operation_type:> operation → ne_list sorts }.
\end{lstlisting}
Given an interpretation of the sorts (mapping each symbolic sort to a carrier), an interpretation for the operations is easily written as an operational type class:
\begin{lstlisting}
  Variables (σ: Signature) (carriers: sorts σ → Type).

@@

What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the `variety'). Since we need a type for the objects in the \lstinline|Category|, at this point we have no choice but to bundle components and proof together in a record:
\begin{lstlisting}
  Variable et: EquationalTheory.

@@

\lstinline|EquationalTheory|'s for semirings and rings are easy to define, and so from section~\ref{varieties} we get corresponding categories in which we can postulate initial objects:
\begin{lstlisting}
  Class Naturals (A: ObjectInVariety semiring_theory) `{!InitialArrow A}: Prop :=
    { naturals_initial:> Initial A }.
\end{lstlisting}
While succinct, this definition is not a satisfactory abstraction because the use of \lstinline|ObjectInVariety| for the type of the \lstinline|A| component `leaks' the fact that we used this one particular universal algebraic construction of the category, which is just an implementation choice. Furthermore, this definition needs an additional layer of class instances to relate it to the \lstinline|SemiRing| class from our algebraic hierarchy.

@@

Some theory and programs about naturals are more easily proved, respectively defined, for concrete implementations and then lifted to the abstract interface. For relatively simple things like decidability of equality, all that is needed to make this lifting easy is a handful of simple lemmas translating the initiality property back into statements saying that \lstinline|naturals_to_semiring| is extensionally equal to other semiring morphisms and involutary between \lstinline|Naturals| implementations. For the example of decidability, these let us define decidability for an arbitrary \lstinline|Naturals| implementation simply by mapping to the \lstinline|nat| implementation using \lstinline|naturals_to_semiring|, deciding equality there, and reflecting the equality or inequality proof back to the abstract implementation using the fact that \lstinline|naturals_to_semiring| is an involution.

@@

Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this quoting. In~\cite{Hints}, Asperti et. al. show that unification hints, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.

@@

With these, we can define the primary ingredient, the \lstinline|Quote| class:
\begin{lstlisting}
  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V'): Type :=
    { quote: Expr (V + V')
    ; eval_quote: eval (merge l r) quote = n }.
\end{lstlisting}
We can think of \lstinline|Quote| as the type for a family of Prolog-like syntax-directed resolution functions, which will take as input \lstinline|V| and \lstinline|l| representing previously encountered holes (opaque subexpressions that could not be destructured further) and their values, along with a concrete term \lstinline|n| to be quoted. Their `output' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values. Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting \lstinline|y| with existing heap \lstinline|x| generates new heap \lstinline|z|''.

@@

  Class Extend := extend: Π {x y} `{SemiGroupOp y} `{MonoidUnit y},
    (x → y) → (seq x → y).
  Class Singleton := singleton: Π x, x → seq x.
\end{lstlisting}
With these, we can define the predicate class for sequences:
\begin{lstlisting}
  Class Sequence
   `{Π a, MonoidUnit (seq a)} `{Π a, SemiGroupOp (seq a)}
   `{Π a, Equiv a → Equiv (seq a)} `{Singleton} `{Extend}: Prop := ...
\end{lstlisting}
%    { ... structural and propriety proofs sufficient to bundle up the categories and arrows ...
%    ; ... AltAdjunction stated using the bundled ... }.
On top of this interface we can build theory about typical list operations such as maps, folds, their relation to \lstinline|singleton| and \lstinline|extend|, et cetera. We can also generically define `bigops' for summation and products, and easily show properties like distributivity, all without ever mentioning cons lists.
